Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-25 paper/paper.qmd
update paper

library(tidyverse)
library(arrow)
library(knitr)
```

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-results, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research.
[Appendix- @sec-data-details] provides external data detail, [Appendix-
@sec-model-details] provides model detail, and [Appendix-@sec-survey] provides an detailed introduction of surveys, sampling and
observational data in our paper.

@@

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
leveraging comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing public
health strategies and legislative decisions.

@@

library(usmap)
library(ggplot2)
library(dplyr)

@@

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

library(ggnewscale)
death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
death_full <- us_map|>
   full_join(death, c("region"="state"))

@@

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates. \# Model

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
second_model <-
  readRDS(file = here::here("models/second_model.rds"))
```

@@

In this analysis, we will use non-probability Respondent-driven sampling to finding respondents. Respondent-driven sampling (RDS), developed by @heckathorn1997, is ideal for studying hidden populations lacking a sampling frame and facing potential stigma or harm. Unlike snowball sampling, RDS offers dual compensation for participation and recruitment, with recruitment occurring peer-to-peer via social networks. This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics, such as abortion, by leveraging trust within communities.

@@

Therefore, the sample frame of our study will be mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision and are reachable and can be contacted by Abortion Rights Advocacy Groups.

@@

We plan to survey 10 respondents for this study. Our sample will consist of mothers who fit the defined sample frame—those who sought but were unable to access abortion services due to restrictive laws, experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups. These mothers must also be willing to participate in and answer our survey questions, ensuring the reliability and depth of the collected data.

@@

>>>>

library(tidyverse)
library(arrow)
library(knitr)
library(usmap)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(rstanarm)
```

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-results, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research.
[Appendix- @sec-data-details] provides external data detail, [Appendix-
@sec-model-details] provides model detail, and [Appendix- @sec-survey]
provides an detailed introduction of surveys, sampling and observational
data in our paper.

@@

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
@tidyverse was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` @arrow efficiently managed parquet
files for larger datasets. We used `modelsummary` package @modelsummary
to generate clean and interpretable summaries of our regression
models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
clean tables and reproducible reports, and `usmap` for geographical
visualizations. We use `here` @here package simplifies the management of
file paths, ensuring a reproducible workflow by creating a consistent
point of reference for file locations, regardless of the operating
system or working directory.Data cleaning was facilitated by `janitor`
@janitor, while `lubridate` @lubridate handled date-time variables.
Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
regression modeling was conducted using `rstanarm` @rstanarm. Together,
these packages supported an efficient, reproducible workflow for
analyzing and presenting our findings.

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
utilizing comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing public
health strategies and legislative decisions.

@@

@@

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
#| 
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
death_full <- us_map|>
   full_join(death, c("region"="state"))

@@

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
first_model <-
  readRDS(file = here::here("models/first_model.rds"))
second_model <-
  readRDS(file = here::here("models/second_model.rds"))
```

@@

In this analysis, we will use non-probability Respondent-driven sampling
to finding respondents. Respondent-driven sampling (RDS), developed by
@heckathorn1997, is ideal for studying hidden populations lacking a
sampling frame and facing potential stigma or harm. Unlike snowball
sampling, RDS offers dual compensation for participation and
recruitment, with recruitment occurring peer-to-peer via social
networks. This method maintains confidentiality, reduces bias through
statistical adjustments, and is particularly effective for sensitive
topics, such as abortion, by utilizing trust within communities.

@@

Therefore, the sample frame of our study will be mothers who live in
areas where they are unable to access abortion services due to
restrictive laws and have experienced the death of their infant within
the first year of life in states with abortion bans following the Dobbs
decision and are reachable and can be contacted by Abortion Rights
Advocacy Groups.

@@

We plan to survey 20 respondents for this study. Our sample will consist
of mothers who fit the defined sample frame—those who were unable to
access abortion services due to restrictive laws, experienced the death
of their infant within the first year of life, and are reachable through
Abortion Rights Advocacy Groups. These mothers must also be willing to
participate in and answer our survey questions, ensuring the reliability
and depth of the collected data.

## Recruitment of Respondents

To recruit participants for this sensitive study, we will use a Respondent-Driven Sampling (RDS) approach in collaboration with  Abortion Rights Advocacy Groups, that have established relationships with our target population. These organizations will serve as initial “seeds” to identify and reach mothers affected by restrictive abortion laws who meet our study criteria. Outreach materials emphasizing confidentiality and the study’s importance will be shared through trusted channels. As part of the RDS methodology, participants may also invite others from their network who share similar experiences, ensuring the sample expands while maintaining anonymity. Interested participants will undergo a screening process and provide informed consent, with the option to complete the survey anonymously online, ensuring safety and confidentiality. To support participant well-being, we will provide access to mental health resources throughout the process, fostering trust and encouraging meaningful participation.

## Handling Non-response bias

Non-response bias is a concern as it can distort survey results and lead to unrepresentative conclusions about the target population. Since our survey addresses a sensitive topic and takes approximately 10–15 minutes to complete, there is a risk of participant dropout or refusal. To minimize this, we provide a clear overview of the survey’s purpose, confidentiality measures, and estimated completion time, fostering trust and encouraging participation.

## Respondent Validation

To ensure the credibility of our data, we implement a respondent validation process. Participants will undergo an initial screening to confirm they meet the study criteria, such as being unable to access abortion services due to restrictive laws and experiencing the loss of an infant within the first year of life. Responses will be reviewed for completeness and consistency, and duplicate entries will be removed to maintain data integrity. Collaboration with trusted support groups and advocacy organizations further enhances respondent reliability, ensuring that our findings accurately reflect the experiences of the target population.

@@

Abortion, or even considering abortion, carries significant social
stigma. Therefore, our survey requires extremely careful phrasing and
design to minimize this stigma’s impact on the study of abortion,
pregnancy, fertility, and related demographic and health outcomes. In
@lindberg2022, researchers tested innovative methods in a national
survey to address persistent underreporting of abortion in the U.S.,
focusing on question design and placement to reduce sensitivity, social
desirability bias, and perceived intrusiveness. Although none of their
experimental approaches significantly improved abortion reporting
compared to the control condition, their study offers valuable insights
for improving abortion measurement. We aim to adapt our survey design
based on their methodologies to enhance data accuracy and respondent
comfort.

This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias. Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., “Prefer not to say”) to address underreporting in sensitive topics. Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy. Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each sectionof our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.

## Solution to the response bias in our survey

We draw on recommendations from @stantcheva2023 to minimize response biases. Common response biases identified in survey design include moderacybias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB). Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.The detailed definitions of these biases are provided in [Appendix-@sec-definition].

To mitigate bias, we enhance our survey in the following ways, drawing on recommendations
from @stantcheva2023:

- Addressing Extreme/Moderacy Bias: We use a minimum of five response options for scale questions to provide more nuanced choices, reducing the likelihood of respondents defaulting to extreme or middle answers.

- Mitigating Response Order Bias: For nominal questions, we randomize response options.

- Minimizing Social Desirability Bias (SDB): Our online survey is carefully designed to minimize social desirability bias (SDB) while addressing the sensitivity of the topic. The survey begins with a clear introduction explaining its purpose—academic research on the impact of restrictive abortion laws—while emphasizing respondent anonymity and confidentiality. Before sensitive questions, participants are reassured that their responses are entirely confidential and will only be used for research purposes. The anonymous online format provides a safe space for participants to share their experiences without fear of judgment or stigma. Additionally, a feedback section at the end invites respondents to voice any concerns or share additional thoughts, fostering trust and enhancing the quality of the data collected.

### Survey Link

The survey has been implemented using Google Forms. You can access it
here: [Survey Link](https://forms.gle/YRrb6Z6EyorHPkuF8).

## Copy of Survey on Restrictive Abortion Laws and Maternal Health

Welcome Section

Introduction: Welcome to our study on the impact of restrictive abortion
laws on maternal and infant health. Your participation in this survey
will help us understand the psychological, social, and systemic impacts
of these laws. Rest assured that your responses are anonymous and will
only be used for academic research purposes.

This survey is conducted by nonpartisan researchers in public health and
social sciences. It consists of 17 carefully designed questions and
should take approximately 10–15 minutes to complete.

Please answer the questions honestly. If you experience any discomfort
while completing the survey, you may stop at any time. For support, we
provide a list of mental health resources at the end of the survey.

Contact Information: Diana Shen Email: diana.shen\@mail.utoronto.ca

Section 1: Demographics and Background Information

1.  What is your age?
    -   Under 18
    -   18–24
    -   25–34
    -   35–44
    -   45–54
    -   55+
    -   Prefer not to say
2.  What is your highest level of education?
    -   Less than high school
    -   High school graduate or equivalent
    -   Some college
    -   Bachelor’s degree
    -   Graduate or professional degree
    -   Prefer not to say

3.What is your marital status? - Single - Married - Divorced - Widowed -
Prefer not to say

Section 2: Understanding Abortion Experiences

Introduction: This section focuses on understanding abortion
experiences, including the circumstances and decisions surrounding them.
Your responses are invaluable in helping researchers and policymakers
improve health services and support for women and families. Please know
that your answers are entirely confidential and will only be used for
research purposes. If you are comfortable, we encourage you to answer as
honestly as possible. If you prefer not to answer, you are welcome to
skip this section.

1.  Are you willing to answer this part?
    -   Yes
    -   No \[Jump to Section 3\]
2.  Did you seek abortion services during your pregnancy?
    -   Yes
    -   No
    -   Prefer not to say
3.  If yes, were you unable to access abortion services due to legal
    restrictions in your state?
    -   Yes
    -   No
    -   Prefer not to say
4.  How did the inability to access abortion services impact your mental
    health during pregnancy?
    -   No impact
    -   Mild impact
    -   Moderate impact
    -   Severe impact
    -   Prefer not to say
5.  Did you receive any support from healthcare providers or community
    organizations during your pregnancy?
    -   Yes
    -   No
6.  If you were unable to access abortion services, what barriers did
    you encounter? (Select all that apply)
    -   Legal restrictions
    -   Financial constraints
    -   Lack of healthcare providers
    -   Distance to clinic
    -   Fear of stigma or judgment
    -   Prefer not to say
7.  If you sought an abortion but were unable to access one, how did
    this affect your mental health during pregnancy?
    -   Increased stress
    -   Anxiety
    -   Depression
    -   Anger or frustration
    -   Feeling of helplessness
    -   Prefer not to say

Section 3: Experiences and Support for Women Facing Pregnancy Challenges

Introduction: This section seeks to understand the experiences and
outcomes of women who, due to restrictive abortion laws, were unable to
access abortion services and subsequently faced the loss of their child.
We recognize that discussing past pregnancies can be sensitive,
especially those involving circumstances such as abortion or the loss of
a child. Your insights are invaluable in identifying areas where
healthcare and support services can be improved. If you prefer not to
answer, you are welcome to skip this section.

1.  Are you willing to answer this part?
    -   Yes
    -   No \[Jump to Section 4\]
2.  Did you experience the loss of your child within the first year of
    life?
    -   Yes
    -   No \[Jump to Section 4\]
    -   Prefer not to say
3.  What was the primary cause of your child’s death as communicated by
    healthcare providers?
    -   Premature birth or related complications
    -   Congenital abnormalities or genetic conditions
    -   Sudden Infant Death Syndrome (SIDS)
    -   Infections (e.g., pneumonia, sepsis)
    -   Birth trauma or delivery complications
    -   Lack of access to timely medical care
    -   Other (please specify)
4.  To what extent do you believe the instances during your pregnancy
    where mental health challenges prevented you from seeking or
    receiving adequate medical care?
    -   1 (No impact)
    -   2
    -   3
    -   4
    -   5 (Significantly)
5.  To what extent do you believe the mental health challenges caused by
    the inability to access abortion services contributed to health
    complications for your child?
    -   Not at all – I don’t believe my mental health challenges had any
        impact on my child’s health.
    -   Slightly – I think there may have been a minor impact on my
        child’s health.
    -   Moderately – I feel my mental health challenges had a noticeable
        impact on my child’s health.
    -   Significantly – I believe my mental health challenges had a
        considerable impact on my child’s health.
    -   Completely – I think my mental health challenges were the
        primary factor in my child’s health complications.
    -   Prefer not to say
6.  Looking back, do you believe access to abortion services could have
    positively affected your mental and physical health during
    pregnancy?
    -   Yes
    -   No
    -   Maybe
    -   Prefer not to say

Section 4: Perspectives and Support

If you feel distressed or need support after completing this survey, the
following resources are available to provide assistance:

1.  Postpartum Support International (PSI)
    -   Website: www.postpartum.net
    -   Helpline: 1-800-944-4773 (Text “Help” to 800-944-4773)
    -   Services: Support for mental health during and after pregnancy,
        including peer support and counseling.
2.  Mental Health America (MHA)
    -   Website: www.mhanational.org
    -   Services: Online screening tools, support networks, and
        educational materials.
3.  SAMHSA National Helpline
    -   Website: www.samhsa.gov
    -   Hotline: 1-800-662-HELP (4357)
    -   Services: Free, confidential referrals for mental health and
        substance use disorders.

Section 5: Feedback

1.  Do you have any concerns or feedback regarding the survey, surveyor,
    or entity?
    -   Your feedback is important to us and will help ensure
        transparency and trust in the research process.

Section 6: Thank You

Thank you for taking the time to complete this survey. Your honest
feedback is invaluable and will help us better understand and address
the experiences of women who have faced similar circumstances. We deeply
appreciate your participation and the courage it takes to share your
experiences.

## Response bias definition {#sec-definition}

In the design of the questionnaire, there will be some common biases that
may occur when running the questionnaire.

@stantcheva2023 defines these biases as:

- **Moderacy response bias**: The tendency to choose mid-scale responses.

- **Extreme response bias**: The inclination to select extreme values on the scale.

- **Response order bias**: Occurs when the position of options influences responses, with the **primacy effect** (favoring early options in written surveys) and **recency effect** (favoring later options in oral surveys) as common forms.

- **Social desirability bias**: Respondents may hide true views to present a socially favorable image, influenced by topic and social context.

- **Acquiescence bias**: The habit of giving positive responses, such as consistently choosing “agree” or “yes.”
\newpage
